---
title: "R Notebook"
output: html_notebook
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview - River Monitoing 22 - DOM Quality 

## Data
- DOM quality indices
- TOC needed
- 5 year average and 2022 data

```{r, results=FALSE, message=FALSE, warning=FALSE}
#install.packages(c("dplyr", "ggplot2", "plyr", "Rmisc", "tidyverse", "lubridate", "stringr"))
Packages <- c("plyr", "dplyr", "ggplot2",  "Rmisc", "tidyverse", "lubridate", "stringr", "tm")
lapply(Packages, library, character.only = TRUE)
```
*Functions* 
The following function will replace all "<LOD"
- har det no å si om det er mellomrm mellom < og tallet?
```{r echo = T, results = 'hide'}
dLimitx <- function(v){
  v <- as.character(v)
  isLimit <- grepl("<", v)  #grepl() is a built-in R function looks for a smaller string of characters
  n <- (sum(isLimit)/(sum(!isLimit)+sum(isLimit)))*100
  v[isLimit] <- as.numeric(gsub("<(.*)", "\\1", v[isLimit]))*(100-n)/100 #gsub(pattern, replacement, x) for substitution
  as.numeric(v)
}

fmt_dcimals <- function(decimals=1){
  function(x) format(x,nsmall = decimals,scientific = FALSE)
}
```

Load the two datafiles of current year and previous years data 
```{r echo = T, results = 'hide'}
Sys.setlocale(locale="no_NO") 
A2021 <- read.table("Hovedelver_UVabs_fra Aquamonitor.txt", header=TRUE, sep="\t", na.string=c(""))
A3yr <- read.table("UVabs data 2017-2020.txt", header=TRUE, sep="\t", na.string=c(""))

C2021 <- read.table("copy_hovedelver_2021.txt", header=TRUE, sep="\t", na.string=c(""))
C3yr <- read.table("DOC_4y.txt", header=TRUE, sep="\t", na.string=c(""))

```

### 1) Create the 5-yr average
- Data 2017-2020 must be combined with 2021
- Both UV-Vis abs and DOC data
- 2022 UV vis data together with DOC

```{r, create the 5-year avaerage}
A3yr1 <- A3yr[ -c(7:8) ]
AA <- rbind(A2021, A3yr1)
```

### 2) Calculate indices
```{r, Calculate indices}

```

### 3) Make summaries for plotting
```{r}

```

Prepare for plotting
```{r}

```

PLOTTING
```{r}

```

